---
title: "Seurat workflow and cluster-specific peak calling"
output: html_notebook
params:
  dataset: "unsorted"
  cellranger_output: "../data/CellRanger/unsorted/"
  result_folder: "../results/Seurat/callpeaks_unsorted/"
  seurat_resolution: 0.8
---

## Summary: 
This notebook provides the steps below:\
1.) Reading and processing h5 CellRanger output\
2.) Seurat workflow\
  - give annotations\
  - normalization\
  - dimension reduction\
  - UMAP visualizations\
3.) Find marker regions \
4.) peaks assay\
  - gene activity assay\
  - Cluster-specific peak calling (CallPeaks function. It depends on MACS2 and linux)\
  
seurat_resolution parameter: \
  by default: 0.8\
  strict: 0.1\



```{r}
suppressPackageStartupMessages({
  library("Seurat", warn.conflicts = FALSE)
  library("Signac", warn.conflicts = FALSE )
  library("glue", warn.conflicts = FALSE)
  library("tidyverse", warn.conflicts = FALSE)
  library("rtracklayer", warn.conflicts = FALSE)
  library("EnsDb.Mmusculus.v79", warn.conflicts = FALSE)
  library("GenomicFeatures", warn.conflicts = FALSE)
  library("GenomeInfoDb", warn.conflicts = FALSE)
  library("cowplot", warn.conflicts = FALSE)
  library("Matrix", warn.conflicts = FALSE)
  # library("DoubletFinder")
})
```


```{r}
set.seed(5)
```

read and process data
```{r}
counts = Read10X_h5(filename = glue("{params$cellranger_output}filtered_peak_bc_matrix.h5"))
metadata = read.csv(
  file = glue("{params$cellranger_output}singlecell.csv"),
  header = TRUE,
  row.names = 1
)

chrom_assay = CreateChromatinAssay(
  counts = counts,
  sep = c(":", "-"),
  genome = "mm10",
  fragments = glue("{params$cellranger_output}fragments.tsv.gz"),
  min.cells = 10,
  min.features = 200
)

```

Seurat workflow and clustering
```{r}
g4 = CreateSeuratObject(counts = chrom_assay,
                        assay = "peaks",
                        meta.data = metadata)

# check peaks in GRanges format
granges(g4)

# give EnsDB mm10 annotation
annotations = suppressWarnings(GetGRangesFromEnsDb(ensdb = EnsDb.Mmusculus.v79, verbose = FALSE))
seqlevelsStyle(annotations) = 'UCSC'
Annotation(g4) = annotations

# normalization
g4 = RunTFIDF(g4)
g4 = FindTopFeatures(g4, min.cutoff = 'q0')
g4 = RunSVD(g4)

```


Non-linear dimension reduction and clustering
```{r}
# Non-linear dimension reduction and clustering
g4 = RunUMAP(object = g4,
             reduction = 'lsi',
             dims = 2:30,
             verbose = FALSE)
g4 = FindNeighbors(object = g4,
                   reduction = 'lsi',
                   dims = 2:30)
g4 = FindClusters(
  object = g4,
  verbose = FALSE,
  resolution = params$seurat_resolution,
  algorithm = 3
)
```

Find all marker regions
```{r}
markers = FindAllMarkers(g4, verbose = FALSE)

top.markers = markers %>% group_by(cluster) %>% top_n(wt = avg_log2FC, n = 5)
DoHeatmap(
  object = g4,
  features = top.markers$gene,
  slot = 'data',
  raster = TRUE
) + scale_fill_gradient(low = "white",  high = "#fc9272")

# export marker regions
markers = markers %>% separate(gene, sep = "-", into = c("chr", "start", "end"), remove = TRUE) %>% 
  dplyr::select("chr", "start", "end", everything()) %>% mutate(start = as.numeric(start), end = as.numeric(end))
write_tsv(markers, glue("{params$result_folder}FindAllMarkers_output.tsv"))
bed = markers %>% dplyr::select("chr", "start", "end")
write_tsv(bed, glue("{params$result_folder}FindAllMarkers_output_bed.tsv"), col_names = FALSE)

```


UMAP visualization by dimension plot
```{r}

# 0.1 returns three clusters
if (params$seurat_resolution == 0.1) {
  cols = c("0" = "#addd8e",
           "1" = "#bdbdbd",
           "2" = "#addd8e")
} else {
  palette = "Set3"
}

if (params$seurat_resolution == 0.1) {
  dim_res0.1 = DimPlot(
    object = g4,
    label = TRUE,
    pt.size = 2,
    label.size = 7,
    repel = TRUE
  ) +
    #NoLegend() +
    #scale_color_brewer(palette = "Set3") +
    xlim(-10, 10) +
    ylim(-10, 10) +
    scale_colour_manual(
      values = cols,
      breaks = c("0", "1"),
      labels = c("oligodendrocytes (GFP+)", "unsorted brain")
    ) +
    ggtitle("unsorted brain G4 scCut&Tag, res = 0.1") +
    theme(
      text = element_text(size = 25),
      plot.title = element_text(size = 20),
      axis.text.x = element_text(size = 25, color = "black"),
      axis.text.y = element_text(size = 25, color = "black")
    )
  dim_res0.1
} else {
  dim = DimPlot(
    object = g4,
    label = TRUE,
    pt.size = 2,
    label.size = 7,
    repel = TRUE
  ) +
    NoLegend() +
    scale_color_brewer(palette = palette) +
    xlim(-10, 10) +
    ylim(-10, 10) +
    ggtitle("unsorted brain G4 scCut&Tag") +
    theme(
      text = element_text(size = 25),
      plot.title = element_text(size = 20),
      axis.text.x = element_text(size = 25, color = "black"),
      axis.text.y = element_text(size = 25, color = "black")
    )
  dim
}



```
```{r}
if (params$seurat_resolution != 0.1) {
  dim = DimPlot(
    object = g4,
    label = TRUE,
    pt.size = 2,
    label.size = 7,
    repel = TRUE
  ) +
    NoLegend() +
    scale_color_brewer(palette = "Set3") +
    xlim(-10, 10) +
    ylim(-10, 10) +
    ggtitle("unsorted brain G4 scCut&Tag") +
    theme(
      text = element_text(size = 25),
      plot.title = element_text(size = 20),
      axis.text.x = element_text(size = 25, color = "black"),
      axis.text.y = element_text(size = 25, color = "black")
    )
  dim
  
  dim_blank = DimPlot(
    object = g4,
    label = TRUE,
    pt.size = 2,
    label.size = 7,
    repel = TRUE
  ) +
    NoLegend() +
    NoAxes() +
    scale_color_brewer(palette = "Set3") +
    xlim(-10, 10) +
    ylim(-10, 10) +
    theme(
      text = element_text(size = 25),
      plot.title = element_text(size = 20),
      axis.text.x = element_text(size = 0, color = "black"),
      axis.text.y = element_text(size = 0, color = "black")
    )
  dim_blank
}
```

save plots
```{r}
if (params$seurat_resolution != 0.1) {
  ggsave(
    glue("{params$result_folder}Seurat_unsorted_UMAP.png"),
    plot = dim,
    width = 10,
    height = 10,
    dpi = 300,
  )
  
  ggsave(
    glue("{params$result_folder}Seurat_unsorted_UMAP.pdf"),
    plot = dim,
    device = "pdf",
    width = 10,
    height = 10,
    dpi = 300,
  )
  
  ggsave(
    glue(
      "{params$result_folder}Seurat_mESC-MEF_UMAP_blankedplot.pdf"
    ),
    plot = dim_blank,
    width = 10,
    height = 10,
    device = "pdf"
  )
} else {
  ggsave(
    glue("{params$result_folder}Seurat_unsorted_UMAP_res0.1.png"),
    plot = dim_res0.1,
    width = 10,
    height = 10,
    dpi = 300,
  )
  
  ggsave(
    glue("{params$result_folder}Seurat_unsorted_UMAP_res0.1.pdf"),
    plot = dim_res0.1,
    width = 10,
    height = 10,
    dpi = 300,
  )
}

```

Create a gene activity matrix
```{r}
gene.activities = GeneActivity(g4)

g4[['GA']] = CreateAssayObject(counts = gene.activities)
g4 = NormalizeData(
  object = g4,
  assay = 'GA',
  normalization.method = 'LogNormalize',
  scale.factor = median(g4$nCount_peaks)
)

```

create GA heatmap
```{r}
ga_markers = FindAllMarkers(g4, only.pos = TRUE, assay = "GA", verbose = FALSE)

ga_top.markers = ga_markers %>% group_by(cluster) %>% top_n(wt = avg_log2FC, n = 5)
DoHeatmap(
  object = g4,
  features = ga_markers$gene,
  slot = 'data',
  raster = TRUE,
  assay = "GA"
) +
  scale_fill_gradient2(low = "white", mid = "white", high = "red") +
  labs(title = " ", fill = "gene act. score") +
  theme(plot.title = element_text(size = 8),
        axis.text.y = element_text(size = 8, color = "black"))

if (params$seurat_resolution == 0.1) {
  ggsave(
    glue("{params$result_folder}GA_heatmap_res0.1.pdf"),
    plot = last_plot(),
    device = "pdf",
    width = 10,
    height = 3
  )
} else {
  ggsave(
    glue("{params$result_folder}GA_heatmap.pdf"),
    plot = last_plot(),
    device = "pdf",
    width = 10,
    height = 3
  )
  
}
```

extract cell barcodes per cluster
```{r}
barcodes = g4@meta.data %>%
  rownames_to_column("barcodes") %>%
  dplyr::select(barcodes, seurat_clusters)

if (params$seurat_resolution == 0.1) {
  write_tsv(barcodes,
            glue("{params$result_folder}barcodes_per_cluster_res0.1.tsv"))
} else {
  write_tsv(barcodes, glue("{params$result_folder}barcodes_per_cluster.tsv"))
}


for (cluster in unique(barcodes$seurat_clusters)) {
  subset = barcodes %>% dplyr::filter(seurat_clusters == cluster)
  if (params$seurat_resolution == 0.1) {
    write_tsv(
      subset,
      glue(
        "{params$result_folder}barcodes_cluster_{as.character(cluster)}_res0.1.tsv"
      ),
      col_names = FALSE
    )
  } else {
    write_tsv(
      subset,
      glue(
        "{params$result_folder}barcodes_cluster_{as.character(cluster)}.tsv"
      ),
      col_names = FALSE
    )
  }
}
```

save Seurat object
```{r}
if (params$seurat_resolution == 0.1) {
  saveRDS(g4, glue("{params$result_folder}unsorted_res0.1.Rds"))
} else {
  saveRDS(g4, glue("{params$result_folder}unsorted.Rds"))
}
```

call cluster-specific peaks by CallPeaks function (MACS2 is needed)
```{r}
peaks = CallPeaks(
  object = g4,
  group.by = "seurat_clusters",
  cleanup = FALSE,
  outdir = params$result_folder,
  effective.genome.size = 2652783500
)

if (params$seurat_resolution == 0.1) {
  write.table(
    as.data.frame(peaks),
    file = glue("{params$result_folder}peaks_per_clusters_res0.1.bed"),
    quote = F,
    sep = "\t",
    row.names = F,
    col.names = F
  )
} else {
  write.table(
    as.data.frame(peaks),
    file = glue("{params$result_folder}peaks_per_clusters.bed"),
    quote = F,
    sep = "\t",
    row.names = F,
    col.names = F
  )
  
}
```
