---
title: "Overlaps with bulk G4 data"
output:
  html_document:
    df_print: paged
  html_notebook: default
  pdf_document: default
params:
  peak_folder: "../results/Seurat/callpeaks_mESC-MEF/peak_sets/"
  bed_folder: "../data/bed/"
  result_folder: "../results/GenomicRanges/mESC-MEF_outputs/"
---

## Summary: \
This notebook provides the steps below:\
1.) Make Venn diagrams between bulk and single-cell datasets \
2.) generate bed files of unique or common peaks between clusters (Seurat resolution 0.1)

```{r}
# packages
suppressPackageStartupMessages({
  library("glue")
  library("tidyverse")
  library("rtracklayer")
  library("data.table")
  library("GenomicFeatures")
  library("ChIPseeker")
  library("ChIPpeakAnno")
  library("Vennerable")
  library("GenomicRanges")
  library("TxDb.Mmusculus.UCSC.mm10.knownGene")
  library("biomaRt")
  library("wigglescout")
  library("GenometriCorr")
})
```

```{r}
create_venn = function(peak1,
                       peak2,
                       type1,
                       type2,
                       filename) {
  # filtered MACS2 peak calls
  peak = fread(glue("{params$peak_folder}{peak1}"))
  peak$type = type1
  peak = GRanges(
    seqnames = peak$V1,
    ranges = IRanges(
      start = peak$V2,
      end = peak$V3,
      names = peak$type,
    )
  )
  
  bulk_peak = fread(glue("{params$bed_folder}{peak2}"))
  bulk_peak$type = type2
  bulk_peak = GRanges(
    seqnames = bulk_peak$V1,
    ranges = IRanges(
      start = bulk_peak$V2,
      end = bulk_peak$V3,
      names = bulk_peak$type,
    )
  )
  
  peak$type = type1
  bulk_peak$type = type2
  gr = c(peak, bulk_peak)
  grl = splitAsList(gr, gr$type)
  grl = unique(grl)
  
  res = makeVennDiagram(Peaks = grl, NameOfPeaks = names(grl))
  
  venn_cnt2venn <- function(venn_cnt) {
    n <- which(colnames(venn_cnt) == "Counts") - 1
    SetNames = colnames(venn_cnt)[1:n]
    Weight = venn_cnt[, "Counts"]
    names(Weight) <- apply(venn_cnt[, 1:n], 1, paste, collapse = "")
    Venn(SetNames = SetNames, Weight = Weight)
  }
  
  pdf(
    file = glue("{params$result_folder}{filename}"),
    # The directory you want to save the file in
    width = 8,
    height = 8
  )
  v <- venn_cnt2venn(res$vennCounts)
  print(plot(v, doWeights = FALSE))
  dev.off()
  
}
```

```{r}
# list of narrowpeaks
narrowpeaks = list.files(params$peak_folder, pattern = "*.narrowPeak")
narrowpeaks
```

Venn with 3 sets - __bulk NPC and cluster0/1 of unsorted scCnT__ \
peak sets: \
cluster 0 and 1: MACS2 peaks following CallPeaks of Seurat \
bulk_peak: MACS2 peaks of bulk NPC CnT
```{r}
# cluster 0 of brain dataset
cluster0_unsorted = fread("../results/Seurat/callpeaks_unsorted/peak_sets/0_peaks_res0.1.narrowPeak")
cluster0_unsorted$type = "cluster0"
cluster0_unsorted = GRanges(
  seqnames = cluster0_unsorted$V1,
  ranges = IRanges(
    start = cluster0_unsorted$V2,
    end = cluster0_unsorted$V3,
    names = cluster0_unsorted$type,
  )
)

# cluster 1 of brain dataset
cluster1_unsorted = fread("../results/Seurat/callpeaks_unsorted/peak_sets/1_peaks_res0.1.narrowPeak")
cluster1_unsorted$type = "cluster1"
cluster1_unsorted = GRanges(
  seqnames = cluster1_unsorted$V1,
  ranges = IRanges(
    start = cluster1_unsorted$V2,
    end = cluster1_unsorted$V3,
    names = cluster1_unsorted$type,
  )
)

# bulk NPC
bulk_peak = fread("../data/bed/bulk_CnT_G4_NPC_mm10.bed")
bulk_peak$type = "NPC_bulk_CnT"
bulk_peak = GRanges(
  seqnames = bulk_peak$V1,
  ranges = IRanges(
    start = bulk_peak$V2,
    end = bulk_peak$V3,
    names = bulk_peak$type,
  )
)

# making Venn
cluster0_unsorted$type = "cluster_0"
cluster1_unsorted$type = "cluster_1"
bulk_peak$type = "NPC_bulk_CnT"
gr = c(cluster0_unsorted, cluster1_unsorted, bulk_peak)
grl = splitAsList(gr, gr$type)
grl = unique(grl)

res = makeVennDiagram(Peaks = grl, NameOfPeaks = names(grl))

venn_cnt2venn <- function(venn_cnt) {
  n <- which(colnames(venn_cnt) == "Counts") - 1
  SetNames = colnames(venn_cnt)[1:n]
  Weight = venn_cnt[, "Counts"]
  names(Weight) <- apply(venn_cnt[, 1:n], 1, paste, collapse = "")
  Venn(SetNames = SetNames, Weight = Weight)
}

pdf(
  file = glue("{params$result_folder}bulk_NPC_MEF-MESCcluster0_1.pdf"),
  # The directory you want to save the file in
  width = 8,
  height = 8
)
v <- venn_cnt2venn(res$vennCounts)
print(plot(v, doWeights = FALSE))
dev.off()
```
Venn with 3 sets - __bulk MEF and cluster0/1 of mESC-MEF scCnT__\

peak sets: \
cluster 0 and 1: MACS2 peaks following CallPeaks of Seurat \
bulk_peak: MACS2 peaks of bulk mESC CnT
```{r}
# cluster 0 of mESC-MEF dataset
cluster0 = fread("../results/Seurat/callpeaks_mESC-MEF/peak_sets/0_peaks_res0.1.narrowPeak")
cluster0$type = "cluster0"
cluster0 = GRanges(
  seqnames = cluster0$V1,
  ranges = IRanges(
    start = cluster0$V2,
    end = cluster0$V3,
    names = cluster0$type,
  )
)

# cluster 1 of mESC-MEF dataset
cluster1 = fread("../results/Seurat/callpeaks_mESC-MEF/peak_sets/1_peaks_res0.1.narrowPeak")
cluster1$type = "cluster1"
cluster1 = GRanges(
  seqnames = cluster1$V1,
  ranges = IRanges(
    start = cluster1$V2,
    end = cluster1$V3,
    names = cluster1$type,
  )
)

# bulk MEF
bulk_peak = fread("../data/bed/bulk_G4_CnT_MEF_rep1_R1.mLb.clN_peaks.broadPeak")
bulk_peak$type = "MEF_bulk_CnT"
bulk_peak = GRanges(
  seqnames = bulk_peak$V1,
  ranges = IRanges(
    start = bulk_peak$V2,
    end = bulk_peak$V3,
    names = bulk_peak$type,
  )
)

# making Venn
cluster0$type = "cluster_0"
cluster1$type = "cluster_1"
bulk_peak$type = "MEF_bulk_CnT"
gr = c(cluster0, cluster1, bulk_peak)
grl = splitAsList(gr, gr$type)
grl = unique(grl)

res = makeVennDiagram(Peaks = grl, NameOfPeaks = names(grl))

venn_cnt2venn <- function(venn_cnt) {
  n <- which(colnames(venn_cnt) == "Counts") - 1
  SetNames = colnames(venn_cnt)[1:n]
  Weight = venn_cnt[, "Counts"]
  names(Weight) <- apply(venn_cnt[, 1:n], 1, paste, collapse = "")
  Venn(SetNames = SetNames, Weight = Weight)
}

pdf(
  file = glue("{params$result_folder}bulk_mESC_MEF-MEFcluster0_1.pdf"),
  # The directory you want to save the file in
  width = 8,
  height = 8
)
v <- venn_cnt2venn(res$vennCounts)
print(plot(v, doWeights = FALSE))
dev.off()

```
Venn with 3 sets - __bulk mESC and cluster0/1 of mESC-MEF scCnT__ \
peak sets: \
cluster 0 and 1: MACS2 peaks following CallPeaks of Seurat \
bulk_peak: MACS2 peaks of bulk mESC CnT
```{r}
# Venn with 3 sets - bulk mESC and cluster0/1 of mESC-MEF scCnT
cluster0 = fread("../results/Seurat/callpeaks_mESC-MEF/peak_sets/0_peaks_res0.1.narrowPeak")
cluster0$type = "cluster0"
cluster0 = GRanges(
  seqnames = cluster0$V1,
  ranges = IRanges(
    start = cluster0$V2,
    end = cluster0$V3,
    names = cluster0$type,
  )
)

cluster1 = fread("../results/Seurat/callpeaks_mESC-MEF/peak_sets/1_peaks_res0.1.narrowPeak")
cluster1$type = "cluster1"
cluster1 = GRanges(
  seqnames = cluster1$V1,
  ranges = IRanges(
    start = cluster1$V2,
    end = cluster1$V3,
    names = cluster1$type,
  )
)

bulk_peak = fread("../data/bed/bulk_CnT_G4_mES_mm10.bed")
bulk_peak$type = "mESC_bulk_CnT"
bulk_peak = GRanges(
  seqnames = bulk_peak$V1,
  ranges = IRanges(
    start = bulk_peak$V2,
    end = bulk_peak$V3,
    names = bulk_peak$type,
  )
)

cluster0$type = "cluster_0"
cluster1$type = "cluster_1"
bulk_peak$type = "mESC_bulk_CnT"
gr = c(cluster0, cluster1, bulk_peak)
grl = splitAsList(gr, gr$type)
grl = unique(grl)

res = makeVennDiagram(Peaks = grl, NameOfPeaks = names(grl))

venn_cnt2venn <- function(venn_cnt) {
  n <- which(colnames(venn_cnt) == "Counts") - 1
  SetNames = colnames(venn_cnt)[1:n]
  Weight = venn_cnt[, "Counts"]
  names(Weight) <- apply(venn_cnt[, 1:n], 1, paste, collapse = "")
  Venn(SetNames = SetNames, Weight = Weight)
}

pdf(
  file = glue("{params$result_folder}bulk_mESC_MEF-MESCcluster0_1.pdf"),
  # The directory you want to save the file in
  width = 8,
  height = 8
)
v <- venn_cnt2venn(res$vennCounts)
print(plot(v, doWeights = FALSE))
dev.off()
```
__brain scCnT overlaps__ \
Determine unique cluster 0/1 and common peaks by wigglescout. \
Definition of uniqueness: 4-fold difference between the two clusters \
Common peaks: findOverlaps (GenomicRanges) with "any" parameter \

Output bed files are the input of deeptools heatmaps
```{r}
bed_output = "../results/GenomicRanges/unsorted_outputs/bulk_overlaps/"
peak_set = "../results/Seurat/callpeaks_unsorted/peak_sets/"
bigwigs = "../results/Seurat/callpeaks_unsorted/cluster_bigwigs/"
list.files(bigwigs)

# scCnT data
cluster0 = fread(glue("{peak_set}0_peaks_res0.1.narrowPeak"))

cluster0_gr = GRanges(
  seqnames = cluster0$V1,
  ranges = IRanges(
    start = cluster0$V2,
    end = cluster0$V3,
    signalValue = cluster0$V7
  )
)

cluster1 = fread(glue("{peak_set}1_peaks_res0.1.narrowPeak"))

cluster1_gr = GRanges(
  seqnames = cluster1$V1,
  ranges = IRanges(
    start = cluster1$V2,
    end = cluster1$V3,
    signalValue = cluster1$V7
  )
)

all = rbind(cluster0, cluster1)
all = all %>% mutate(type = "unsorted")

all = GRanges(
  seqnames = all$V1,
  ranges = IRanges(
    start = all$V2,
    end = all$V3,
    signalValue = all$V7
  )
)

read_cov = bw_loci(glue(bigwigs, "0_res0.1_oligo.bw"), glue(bigwigs, "1_res0.1_brain_cells.bw") , loci = all)
read_cov = as.data.frame(read_cov)
read_cov = inner_join(read_cov, as.data.frame(all), by = c("seqnames", "start", "end"))

# only cluster 0 peaks (4-fold difference between cluster 0 and 1)
only_cluster0 = read_cov %>% dplyr::filter(!X0_res0.1_oligo == Inf) %>% 
  dplyr::filter(X0_res0.1_oligo >= 4) %>% 
  dplyr::select(seqnames, start, end, signalValue)

write_tsv(only_cluster0, glue("{peak_set}0_unique_peaks_res0.1.tsv"))

only_cluster0 = only_cluster0 %>% dplyr::select(seqnames, start, end)

write.table(
  only_cluster0,
  glue("{bed_output}MACS2_cl0_only.bed"),
  quote = FALSE,
  sep = "\t",
  row.names = FALSE,
  col.names = FALSE
)

read_cov = bw_loci(glue(bigwigs, "1_res0.1_brain_cells.bw"), glue(bigwigs, "0_res0.1_oligo.bw") , loci = all)
read_cov = as.data.frame(read_cov)
read_cov = inner_join(read_cov, as.data.frame(all), by = c("seqnames", "start", "end"))

only_cluster1 = read_cov %>% dplyr::filter(!X1_res0.1_brain_cells == Inf) %>% 
  dplyr::filter(X1_res0.1_brain_cells >= 4) %>% 
  dplyr::select(seqnames, start, end, signalValue)

write_tsv(only_cluster1, glue("{peak_set}1_unique_peaks_res0.1.tsv"))

# only cluster 1 peaks (4-fold difference between cluster 1 and 0)
only_cluster1 = only_cluster1 %>% dplyr::select(seqnames, start, end)

write.table(
  only_cluster1,
  glue("{bed_output}MACS2_cl1_only.bed"),
  quote = FALSE,
  sep = "\t",
  row.names = FALSE,
  col.names = FALSE
)

## overlap between cluster 0,1
ol = findOverlaps(cluster0_gr, cluster1_gr, type = "any", ignore.strand = FALSE)

# intersection
both = cluster0_gr[queryHits(ol)]
both = data.frame(
  seqnames = seqnames(both),
  starts = start(both) - 1,
  ends = end(both)
)

write.table(
  both,
  glue("{bed_output}MACS2_intersection.bed"),
  quote = FALSE,
  sep = "\t",
  row.names = FALSE,
  col.names = FALSE
)
```

__mESC-MEF scCnT overlaps__ \
Determine unique cluster 0/1 and common peaks by wigglescout. \
Definition of uniqueness: 4-fold difference between the two clusters \
Common peaks: findOverlaps (GenomicRanges) with "any" parameter \

Output bed files are the input of deeptools heatmaps 
```{r}
bed_output = "../results/GenomicRanges/mESC-MEF_outputs/bulk_overlaps/"
peak_set = "../results/Seurat/callpeaks_mESC-MEF/peak_sets/"
bigwigs = "../results/Seurat/callpeaks_mESC-MEF/cluster_bigwigs/"
list.files(bigwigs)

# scCnT data
cluster0 = fread(glue("{peak_set}0_peaks_res0.1.narrowPeak"))

cluster0_gr = GRanges(
  seqnames = cluster0$V1,
  ranges = IRanges(
    start = cluster0$V2,
    end = cluster0$V3,
    signalValue = cluster0$V7
  )
)

cluster1 = fread(glue("{peak_set}1_peaks_res0.1.narrowPeak"))

cluster1_gr = GRanges(
  seqnames = cluster1$V1,
  ranges = IRanges(
    start = cluster1$V2,
    end = cluster1$V3,
    signalValue = cluster1$V7
  )
)

all = rbind(cluster0, cluster1)
all = all %>% mutate(type = "unsorted")

all = GRanges(
  seqnames = all$V1,
  ranges = IRanges(
    start = all$V2,
    end = all$V3,
    signalValue = all$V7
  )
)

read_cov = bw_loci(glue(bigwigs, "0_res0.1.bigwig"), glue(bigwigs, "1_res0.1.bigwig") , loci = all)
read_cov = as.data.frame(read_cov)
read_cov = inner_join(read_cov, as.data.frame(all), by = c("seqnames", "start", "end"))

# only cluster 0 peaks (4-fold difference between cluster 0 and 1)
only_cluster0 = read_cov %>% dplyr::filter(!X0_res0.1 == Inf) %>% 
  dplyr::filter(X0_res0.1 >= 4) %>% 
  dplyr::select(seqnames, start, end, signalValue)

write_tsv(only_cluster0, glue("{peak_set}0_unique_peaks_res0.1.tsv"))

only_cluster0 = only_cluster0 %>% dplyr::select(seqnames, start, end)

write.table(
  only_cluster0,
  glue("{bed_output}MACS2_cl0_only.bed"),
  quote = FALSE,
  sep = "\t",
  row.names = FALSE,
  col.names = FALSE
)

read_cov = bw_loci(glue(bigwigs, "1_res0.1.bigwig"), glue(bigwigs, "0_res0.1.bigwig") , loci = all)
read_cov = as.data.frame(read_cov)
read_cov = inner_join(read_cov, as.data.frame(all), by = c("seqnames", "start", "end"))

only_cluster1 = read_cov %>% dplyr::filter(!X1_res0.1 == Inf) %>% 
  dplyr::filter(X1_res0.1 >= 4) %>% 
  dplyr::select(seqnames, start, end, signalValue)

write_tsv(only_cluster1, glue("{peak_set}1_unique_peaks_res0.1.tsv"))

# only cluster 1 peaks (4-fold difference between cluster 1 and 0)
only_cluster1 = only_cluster1 %>% dplyr::select(seqnames, start, end)

write.table(
  only_cluster1,
  glue("{bed_output}MACS2_cl1_only.bed"),
  quote = FALSE,
  sep = "\t",
  row.names = FALSE,
  col.names = FALSE
)

## overlap between cluster 0,1
ol = findOverlaps(cluster0_gr, cluster1_gr, type = "any", ignore.strand = FALSE)

# intersection
both = cluster0_gr[queryHits(ol)]
both = data.frame(
  seqnames = seqnames(both),
  starts = start(both) - 1,
  ends = end(both)
)

write.table(
  both,
  glue("{bed_output}MACS2_intersection.bed"),
  quote = FALSE,
  sep = "\t",
  row.names = FALSE,
  col.names = FALSE
)
```
